Computer-Assisted Analysis of the \\ Anderson-Hájek Ontological Controversy

### 2017-01-02 Papers/2017/LogicaUniversalis-HajekAndersonControversy/main.tex
modifications requested by report 2

\noindent The axioms in G\"odel's ontological proof
\citep{GoedelNotes,ScottNotes} (cf. Fig. \ref{fig:Goedel}) entail what is called \emph{modal collapse}
\citep{Sobel1987,SobelBook2004}: the formula $\varphi \rightarrow \Box
\varphi$, abbreviated as MC, holds for any formula $\varphi$ and not
just for $\exists x. \mathit{God}(x)$ as intended.

@@

This fact, which
has recently been confirmed with higher-order automated theorem
provers \citep{C40,J30}, has led to strong criticism of the argument
and stimulated attempts to remedy the problem.
\citet{Hajek_Magari_and_others_1996,Hajek_der_Mathematiker_2001}
proposed the use of cautious instead of full comprehension principles,
and \citet{fitting02:_types_tableaus_god} took
greater care of the semantics of higher-order quantifiers
in the presence of modalities. Others, such as
\citet{anderson90:_some_emend_of_goedel_ontol_proof}, \citet{Hajek2002}
and \citet{bjordal99}, proposed emendations of G\"odel's
axioms and definitions. They require neither comprehension
restrictions nor more complex semantics. Therefore, they are
technically simpler to analyze with computer support. We have
formalized those using the proof assistant Isabelle/HOL \citep{Isabelle}
together with the automated higher-order reasoners Leo-II \citep{C26},
Satallax \citep{brown2012satallax}, Metis
\citep{Hurd03first-orderproof}, and Nitpick \citep{Nitpick}.   Our
formalizations\footnote{The   formalizations are available in the
subdirectories \url{Anderson}, \url{Hajek} and \url{Bjordal} at \url{gi
thub.com/FormalTheology/GoedelGod/blob/master/Formalizations/Isabelle/
}.} employ the embedding of higher-order modal logic (HOML) in
classical higher-order logic (HOL) as introduced in previous work
\citep{C40,J30,J23}. We explored the effect of different domain
conditions on the provability of lemmas, theorems and even axioms.
This was motivated by a controversy between Hájek and Anderson
regarding the redundancy of some axioms in Anderson's emendation. In
\emph{constant domain semantics}, the individual domains are the same
in all possible worlds. In \emph{varying domain semantics}, the
domains may vary from world to world. This variation is technically
encoded with the help of an existence relation expressing which
individuals actually exist in each world. Quantifiers are then
uniformly formalized as \emph{actualistic quantifiers} (i.e. guarded by
the existence relation). Our main results are summarized here.

For all emendations and variants discussed here, the axioms and
definitions have been shown to be consistent and not to entail modal
collapse.

\begin{samepage}
The following definitions of redundancy, superfluousness and independence are used throughout this paper:

\begin{definition}
An axiom $A$ is \textbf{redundant} w.r.t. a set of axioms $S$ iff $S$ entails $A$. 
\end{definition}

\begin{definition}
An axiom $A$ is \textbf{superfluous} w.r.t. axiom set $S$ iff $S \setminus \{ A \}$ entails T3.
\end{definition}

\begin{definition}
  An axiom $A$ is \textbf{independent} of a set of axioms $S$ iff
  there are models of $S$ where $A$ is true and other models of
  $S$ where $A$ if false.
\end{definition}
\end{samepage}

\clearpage

For both \textbf{constant domain semantics} and \textbf{varying domain
semantics}, the following results hold for \emph{Anderson's
Emendation} (cf. Fig. \ref{fig:Anderson}): T1, C
and T3$'$ can be quickly automated (in logics \K, \K and \KB,
respectively); the axioms A4 and A5$'$ are proven redundant (the former in logic \KFourB and the latter
already in \K); a trivial countermodel (with two worlds and  two
individuals) for MC is generated by Nitpick (for all mentioned
logics); all axioms and definitions are shown to be mutually
consistent.

@@

The redundancy of A4 and A5 is particularly controversial. Magari
\cite{Magari1988} claimed that A4 and A5 are superfluous, arguing that T3 is true in all
models of the other axioms and definitions by Gödel.
\citet[pp.~5--6]{Hajek_Magari_and_others_1996} investigated this
further, and claimed that Magari's claim is not valid, but is
nevertheless true  under additional silent assumptions by Magari.
Moreover, \citet[p.~2]{Hajek_Magari_and_others_1996} cites his earlier
work\footnote{Although \citep{Hajek_der_Mathematiker_2001} precedes
\citep{Hajek_Magari_and_others_1996} in writing, it was published only
5 years later, in German.} \citep{Hajek_der_Mathematiker_2001}, where
he claims (in Theorem 5.3) that for Anderson's emended theory
\citep{anderson90:_some_emend_of_goedel_ontol_proof}, A4 and A5 are
not only superfluous, but also redundant. \citet[footnote 1 in
p.~1]{AndersonGettings}, in a footnote, rebutted Hájek's claim,
arguing that the redundancy of A4 and A5 holds only under constant
domain semantics, while Anderson's emended theory ought to be taken
under Cocchiarella's semantics \citep{Cocchiarella} (a varying domain
semantics). Our results show that Hájek was originally right, under
both constant and varying domain semantics.

@@

\begin{figure}
\begin{sideways}
\includegraphics[width=1.2\textwidth]{Summary.png} (to be removed)
\end{sideways}
%
\begin{sideways}
\scalebox{.9}{
\begin{tabular}{lccccccccccccc} 
\toprule
Investigated Variant & D1$'$ & A:A1$'$ & A2$'$ & A3$'$ & A4 & A4$'$ & H:A4 & A5 & A5$'$
  & H:A5 & T3 & T3$'$ & MC \\  
\midrule 
Scott (const) & - & - & - & - & \tabboxs{\tabred}{N/I} & - & - & \tabboxs{\tabred}{N/I} & - & - & P & -  & P \\
Scott (var) & - & - & - & - & \tabboxs{\tabred}{N/I} & - & - & \tabboxs{\tabred}{N/I} & - & - & P & -  & P \\
Anderson (const) & - & - & - & - & \tabbox{\tabred}{R(K4B)} & - & - & - & \tabboxs{\tabred}{R}  & - & - & P  & CS \\
Anderson (var) & - & - & - & - & \tabbox{\tabred}{R(K4B)} & - & - & - & \tabboxs{\tabred}{R}  & - & - & P  & CS \\
Anderson (mix) & - & - & - & - & \tabbox{\tabred}{R(K4B)} & - & - & - & \tabboxs{\tabred}{I}  & - & - & \tabboxs{\tabred}{CS} & CS \\
H\'ajek AOE$'$ (var) & - & - & \tabboxs{\tabyellow}{CS}  &  & \tabboxs{\tabred}{S/I} & - & - & - & \tabboxs{\tabred}{S/I}  & - & - & P(KB) & CS \\
H\'ajek AOE$'$\_0 (var) & - & - & - & \tabboxs{\tabyellow}{CS} & \tabboxs{\tabgreen}{R} & - & - & - & S/U & - & - & P(KB) & CS \\
H\'ajek AOE$''$ (var) & - & - &  &  & - & - & \tabboxs{\tabred}{S/I} & - & - & \tabboxs{\tabred}{S/I} & - & P(KB) & CS \\
Anderson (simp) (var) & - & R & R  &  &  & R(K4B) & - & - & & - & - &  & \\
Bjørdal (const) & \tabbox{\tabgreen}{R(K4)} & - & \tabboxs{\tabgreen}{R} & \tabboxs{\tabgreen}{R} & - & \tabbox{\tabgreen}{R(KT)} & - & - &\tabboxs{\tabyellow}{N/I} & - & - & \tabbox{\tabgreen}{P(KB)} &  CS \\
Bjørdal (var) & \tabboxs{\tabyellow}{CS} & - & \tabboxs{\tabgreen}{R} & \tabboxs{\tabgreen}{R} & - & \tabbox{\tabgreen}{R(KT)} & - & - &\tabboxs{\tabyellow}{N/I} & - & - & \tabbox{\tabgreen}{P(KB)} &  CS \\
\bottomrule
\end{tabular}
}
\end{sideways}
\caption{Summary of Results.}
\label{fig:summary}
\end{figure}

@@

Using our approach, the formalization and (partly) automated analysis
of several variants of G\"odel's ontological argument has been
surprisingly straightforward. The provers not only confirmed many
claimed results, but also exposed a few mistakes and novel insights
in a long standing controversy.
We believe the technology employed in
this work is ready to be fruitfully adopted in larger scale by
philosophers.

>>>>

\section{Introduction}
We present an exemplary study in \emph{Computational Metaphysics},
which is an emerging, interdisciplinary field aiming at the rigorous
formalisation and deep logical assessment of rational, philosophical
arguments on the computer.  The particular focus here is on the
ontological argument for the existence of God.  While related work
\citep{oppenheimer11,rushby13} has concentrated on Anselm's simpler original
version of the ontological argument and formalized it in classical
predicate logic, we here focus on modern variants of it requiring
higher-order modal logic. More precisely, the variants studied here are
all emendations of Kurt Gödel's seminal contribution
\citep{GoedelNotes}, respectively, Dana Scott's variant of it
\cite{ScottNotes} (cf. Fig.~\ref{fig:Goedel}).

In preceding experiments \citep{C55,C40,J30}, we have already
demonstrated that the technology we employ is well suited for the
task. In fact, computers may even contribute philosophically relevant
new knowledge. For example, the theorem prover Leo-II \cite{Leo-II}
detected an inconsistency in Gödel's original script
\citep{GoedelNotes} of the argument, which was undetected by
philosophers before.  This inconsistency obviously renders Gödel's
original version of the argument as pointless.  However, as our
preceding experiments also confirmed, Scott's variant
\citep{ScottNotes} from Fig.~\ref{fig:Goedel} is immune to this
issue.\footnote{Gödel avoids the first conjunct $\varphi(x)$ in the
  definition of \emph{essence (D2)}, which Scott added for cosmetic
  reason. However, as Leo-II detected, one can prove from Gödel's
  definition that the empty property becomes an essence of every
  individual, which, together with theorem T1, axiom A5 and the
  defintion of \emph{necessary existence (D3)} causes the
  inconsistency. For more details see \citet{C55}.} Hence, in the
remainder of this article the term \emph{Gödel's ontological argument}
actually refers to Scott's variant of it, whose soundness has been
formally 
verified with our technology (for logics entailing higher-order modal
logic KB with constant domain or varying domain semantics). We here
assume some familiarity with these related experiments and papers.

@@

The particular motivation and context for the work presented in the given
article is as follows: It is well known that the axioms in G\"odel's
ontological proof (cf. Fig. \ref{fig:Goedel}) entail what is called
\emph{modal collapse} \citep{Sobel1987,SobelBook2004}. This means that
the formula $\varphi \rightarrow \Box \varphi$, abbreviated as MC,
holds for any formula $\varphi$ and not just for
$\exists x. \mathit{God}(x)$ as intended.  Hence, contingent truth
implies necessary truth, or, in other words, there are no unnecessary
contingent truths. In fact, one may even go as far and interpret MC as a result
against free will.
%
The fact that MC follows from Gödel's axioms, which has also been
confirmed with higher-order automated theorem provers in our previous
experiments \citep{C40,J30}, has led to strong criticism of the
argument and stimulated attempts to remedy the problem.
\citet{Hajek_Magari_and_others_1996,Hajek_der_Mathematiker_2001}
proposed the use of cautious instead of full comprehension principles,
and \citet{fitting02:_types_tableaus_god} took greater care of the
semantics of higher-order quantifiers in the presence of
modalities. Others, such as
\citet{anderson90:_some_emend_of_goedel_ontol_proof},
\citet{Hajek2002} and \citet{bjordal99}, proposed emendations of
G\"odel's axioms and definitions. They require neither comprehension
restrictions nor more complex semantics and are therefore technically
simpler to analyze with computer support. Hence, we here concentrate
on the latter emendations of Gödel's argument and leave the former
ones for future work.  Again, familiarity with the papers by
\citet{anderson90:_some_emend_of_goedel_ontol_proof},
\citet{Hajek2002} and \citet{bjordal99} is assumed, since we here are only
briefly commenting/assessing various claims made in there.

Our formalizations employ the embedding of higher-order modal logic
(HOML) in classical higher-order logic (HOL) as introduced in previous
work \citep{C40,J23} as enabling technology.  We explored the effect
of different domain conditions on the provability of lemmas, theorems
and even axioms.  This was motivated by a controversy between Hájek
and Anderson regarding the redundancy of some axioms in Anderson's
emendation. In \emph{constant domain semantics} (\emph{possibilist
  notion of quantification}), the individual domains are the same in
all possible worlds. In \emph{varying domain semantics}
(\emph{actualist notion of quantification}), the domains may vary from
world to world. The latter notion is technically encoded in our
approach with the help of an existence relation expressing which
individuals actually exist in each world. In other words, actualist
quantification is formalized as possibilist quantification guarded by
the existence relation.\footnote{EDNOTE: This part was
  misleading/wrong before; I reworded it.} This technical solution
enables us to flexibly switch between these different notions of
quantification and to even mix them. When we mention varying
domain semantics (actualist quantification) below, we in fact mean
varying domain semantics for the domain of individuals only, while for
the other types we still assume the standard constant domain semantics
(possibilist quantification).  This is, to our best knowledge, in line 
with the intention of the authors (it would of course be easily
possible in our approach to study other combinations).

In our
experiments we have utilized the proof assistant Isabelle/HOL
\citep{Isabelle} together with the automated higher-order reasoners
Leo-II \citep{Leo-II}, Satallax \citep{Satallax}, Metis
\citep{Hurd03first-orderproof}, and Nitpick \citep{Nitpick}.

Our main results, which we have
presented at the 1st World Congress on Logic and Religion \citep{C41},
are presented in more detail below. The corresponding formalizations,
containing all the details, are available online: see the
subdirectories \url{Anderson}, \url{Hajek} and \url{Bjordal} at
\url{github.com/FormalTheology/GoedelGod/blob/master/Formalizations/Isabelle/}.
\footnote{Ideally, Logica Universalis should provide means for
  permanent storing of such additional material.}

\section{Results of the Experiments}
For \emph{all emendations and variants} discussed here, the axioms and
definitions have been shown to be consistent and not to entail modal
collapse.

Next, we address some controversially debated claims regarding the redundancy,
superfluousness and independence of particular axioms in the
investigated emendations and variants. Several novel findings are
reported.

%\begin{samepage}
The following definitions of redundancy, superfluousness and independence are used throughout this paper:

\begin{definition}
An axiom $A$ is \emph{redundant} w.r.t. a set of axioms $S$ iff $S$ entails $A$. 
\end{definition}

\begin{definition}
An axiom $A$ is \emph{superfluous} w.r.t. axiom set $S$ iff $S \setminus \{ A \}$ entails T3.
\end{definition}

\begin{definition}
  An axiom $A$ is \emph{independent} of a set of axioms $S$ iff
  there are models of $S$ where $A$ is true and other models of
  $S$ where $A$ if false.
\end{definition}
%\end{samepage}

%\clearpage

For both {constant domain semantics} (possibilist quantification) and {varying domain
semantics} (actualist quantification), the following results hold for \emph{Anderson's
Emendation} (cf. Fig. \ref{fig:Anderson}): T1, C
and T3$'$ can be quickly automated (in logics \K, \K and \KB,
respectively); the axioms A4 and A5$'$ are proven redundant (the former in logic \KFourB and the latter
already in \K); a trivial countermodel (with two worlds and  two
individuals) for MC is generated by Nitpick (for all mentioned
logics); all axioms and definitions are shown to be mutually
consistent.

@@

The redundancy of A4 and A5 is particularly controversial. Magari
\cite{Magari1988} claimed that A4 and A5 are superfluous, arguing that T3 is true in all
models of the other axioms and definitions by Gödel.
\citet[pp.~5--6]{Hajek_Magari_and_others_1996} investigated this
further, and claimed that Magari's claim is not valid, but is
nevertheless true  under additional silent assumptions by Magari.
Moreover, \citet[p.~2]{Hajek_Magari_and_others_1996} cites his earlier
work\footnote{Although \citep{Hajek_der_Mathematiker_2001} precedes
\citep{Hajek_Magari_and_others_1996} in writing, it was published only
5 years later, in German.} \citep{Hajek_der_Mathematiker_2001}, where
he claims (in Theorem 5.3) that for Anderson's emended theory
\citep{anderson90:_some_emend_of_goedel_ontol_proof}, A4 and A5 are
not only superfluous, but also redundant. \citet[footnote 1 in
p.~1]{AndersonGettings} rebutted Hájek's claim,
arguing that the redundancy of A4 and A5 holds only under constant
domain semantics, while Anderson's emended theory ought to be taken
under Cocchiarella's semantics \citep{Cocchiarella} (a varying domain
semantics). Our results show that Hájek was originally right, under
both constant and varying domain semantics (for the domain of individuals).

@@

\begin{figure}
\begin{sideways}
%\includegraphics[width=1.2\textwidth]{Summary.png} \fbox{EDNOTE: to be removed}
\end{sideways}
%
\begin{sideways}
%\scalebox{.9}{
\resizebox{19cm}{4.5cm}{
\begin{tabular}{lccccccccccccc} 
\toprule
Investigated Variant & D1$'$ & A:A1$'$ & A2$'$ & A3$'$ & A4 & A4$'$ & H:A4 & A5 & A5$'$
  & H:A5 & T3 & T3$'$ & MC \\  
\midrule 
Scott (const) & - & - & - & - & \tabboxs{\tabred}{N/I} & - & - & \tabboxs{\tabred}{N/I} & - & - & P & -  & P \\
Scott (var) & - & - & - & - & \tabboxs{\tabred}{N/I} & - & - & \tabboxs{\tabred}{N/I} & - & - & P & -  & P \\
Anderson (const) & - & - & - & - & \tabbox{\tabred}{R(K4B)} & - & - & - & \tabboxs{\tabred}{R}  & - & - & P  & CS \\
Anderson (var) & - & - & - & - & \tabbox{\tabred}{R(K4B)} & - & - & - & \tabboxs{\tabred}{R}  & - & - & P  & CS \\
Anderson (mix) & - & - & - & - & \tabbox{\tabred}{R(K4B)} & - & - & - & \tabboxs{\tabred}{I}  & - & - & \tabboxs{\tabred}{CS} & CS \\
H\'ajek AOE$'$ (var) & - & - & \tabboxs{\tabyellow}{CS}  &  & \tabboxs{\tabred}{S/I} & - & - & - & \tabboxs{\tabred}{S/I}  & - & - & P(KB) & CS \\
H\'ajek AOE$'$\_0 (var) & - & - & - & \tabboxs{\tabyellow}{CS} & \tabboxs{\tabgreen}{R} & - & - & - & S/U & - & - & P(KB) & CS \\
H\'ajek AOE$''$ (var) & - & - &  &  & - & - & \tabboxs{\tabred}{S/I} & - & - & \tabboxs{\tabred}{S/I} & - & P(KB) & CS \\
Anderson (simp) (var) & - & R & R  &  &  & R(K4B) & - & - & & - & - &  & \\
Bjørdal (const) & \tabbox{\tabgreen}{R(K4)} & - & \tabboxs{\tabgreen}{R} & \tabboxs{\tabgreen}{R} & - & \tabbox{\tabgreen}{R(KT)} & - & - &\tabboxs{\tabyellow}{N/I} & - & - & \tabbox{\tabgreen}{P(KB)} &  CS \\
Bjørdal (var) & \tabboxs{\tabyellow}{CS} & - & \tabboxs{\tabgreen}{R} & \tabboxs{\tabgreen}{R} & - & \tabbox{\tabgreen}{R(KT)} & - & - &\tabboxs{\tabyellow}{N/I} & - & - & \tabbox{\tabgreen}{P(KB)} &  CS \\
\bottomrule
\end{tabular}
}
\end{sideways}
\caption{Summary of Results.}
\label{fig:summary}
\end{figure}

@@

\section{Conclusion}
Using our approach, that is, semantical embeddings of variants of higher-order
modal logics in classical higher-order logic, the formalization and
(partly) automated analysis of several variants of G\"odel's
ontological argument has been surprisingly straightforward. The
higher-order provers we employed not only confirmed many claimed results, but also exposed a
few mistakes and novel insights in a long standing controversy.  We
believe the technology employed in this work is ready to be fruitfully
adopted in larger scale by philosophers.  Moreover, our approach is
by no means restricted to metaphysics or philosophy, and further work
includes similar applications in other areas, including, for example,
law, politics and ethics. In fact, we claim, that our approach, at
least to some degree, a realises Leibniz' dream of a characteristica universalis
and calculus ratiocinator.
